---
title: |
  | DS221
  | Data Visualization and Storytelling
subtitle: |
  | --------------------------------------------------
  | PORTFOLIO 1
  | --------------------------------------------------
output: 
  pdf_document: 
    keep_tex: true
header-includes:
  - "\\usepackage{titling}"
  - "\\pretitle{\\begin{center}\\LARGE\\includegraphics[width=4cm]{logo.jpg}\\\\[\\bigskipamount]}"
  - "\\posttitle{\\end{center}}"
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r, include=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com/"))
```

# How to Create a 3D Population Density Map in R

### **1. Installing Required Packages**:

```{r eval=FALSE, include=FALSE}
install.packages("sf", dependencies=TRUE) 
install.packages("tmap", dependencies=TRUE) 
install.packages("mapview", dependencies=TRUE)
install.packages("stars", dependencies=TRUE)
install.packages("rayshader", dependencies=TRUE)
install.packages("MetBrewer", dependencies=TRUE)
install.packages("rayrender")
install.packages("extrafont", dependencies=TRUE)
install.packages("magick", dependencies=TRUE)
install.packages("plotly", dependencies = TRUE)
install.packages("showtext")
```

**install.packages("sf", dependencies=TRUE):**

This line installs the '*sf*' package, which is crucial for working with
spatial data. It provides functions and tools for reading, writing, and
manipulating spatial data formats.

**install.packages("tmap", dependencies=TRUE):**

This line installs the '*tmap*' package, which is used for creating
thematic maps. It offers a simple and intuitive interface for
visualizing spatial data and creating informative maps.

**install.packages("mapview", dependencies=TRUE)**:

This line installs the '*mapview*' package, which provides interactive
viewing of spatial data. It allows us to explore our spatial data
interactively, making it easier to understand and analyze.

**install.packages("stars", dependencies=TRUE)**:

This line installs the '*stars*' package, which is used for
*spatiotemporal* data manipulation and analysis. It provides functions
for handling raster and vector data, as well as tools for visualization
and analysis.

**install.packages("rayshader", dependencies=TRUE):**

This line installs the '*rayshader*' package, which is used for creating
3D visualizations of spatial data. It allows us to generate a realistic
3D terrain map and visualize spatial data in a unique and immersive way.

**install.packages("MetBrewer", dependencies=TRUE**):

This line installs the '*MetBrewer*' package, which provides color
palettes for our thematic mapping. It offers a variety of color schemes
optimized for visualizing meteorological data, making our map more
informative and visually appealing.

**install.packages("rayrender"**):

This line installs the '*rayrender*' package, which is used for creating
high-quality 3D renderings. It will provide functions for generating
photorealistic images and animations, allowing us to visualize spatial
data in a realistic and visually stunning way.

**install.packages("extrafont", dependencies=TRUE):**

This line installs the '*extrafont*' package, which is used for
importing and managing additional fonts. It allows us to use a wide
range of fonts in our plots and graphics, enhancing their visual appeal
and readability.

**install.packages("magick", dependencies=TRUE):** This line installs
the '*magick*' package, which is used for working with images . It
provides functions for reading, writing, and manipulating images, as
well as tools for creating animations and visual effects.

**install.packages("plotly", dependencies = TRUE):** This line installs
the '*plotly*' package, which is used for creating interactive plots and
graphics. It allows us to create interactive visualizations with
features such as zooming, panning, and hovering, making it easier to
explore and understand your data.

These packages are essential for various aspects of spatial data
analysis, visualization, and graphical representation in R. By
installing them, we gain access to a wide range of tools and functions
that streamline our workflow and enhance the quality of our work.

### **2. Load Packages and Set Options**:

Load the required libraries and set the RGL options:

```{r}
# Set option to use rgl graphics device
options(rgl.useNULL = FALSE)

# Load required packages for data manipulation and visualization
# Load the tidyverse package, which includes many useful packages for data manipulation and visualization
require(tidyverse)

# Load the sf package for working with spatial data
require(sf)

# Load the tmap package for creating thematic maps
require(tmap)

# Load the ggplot2 package for creating versatile plots
require(ggplot2)

# Load the mapview package for interactive viewing of spatial data
require(mapview)

# Load the stars package for spatiotemporal data manipulation and analysis
require(stars)

# Load the rayshader package for creating 3D visualizations
require(rayshader)

# Load the MetBrewer package for color palettes optimized for thematic mapping
require(MetBrewer)

# Load the colorspace package for manipulating and selecting colors
require(colorspace)

# Load the rayrender package for creating high-quality 3D renderings
require(rayrender)

# Load the magick package for working with images
require(magick)

# Load the extrafont package for importing and managing additional fonts in plots
require(extrafont)

# Importing the plotly library for creating interactive plots
require(plotly)
```

Notice that **require()** was used for this. R then checks if the
packages are installed already.It is the preferred method for loading
packages in R due to its automatic installation, error handling,
namespace management, conditional loading capabilities, and simplicity.
It ensures that your code runs smoothly and reliably, even when working
with multiple packages and dependencies.

### **3. Load and Transform Data**:

You'll need to load the population data and administrative boundaries
for the chosen country (*with this work, the country is Cambodia*),
transforming them into a suitable coordinate system. The data is
downloaded from [**Kontur Population.**]{.underline}

```{r}
# Read population data for Cambodia from a GeoPackage file and transform it to a specific CRS
Cambodia_hex <- st_read("kontur_KHpopulation_KH_20231101.gpkg") %>% st_transform(3106)

# Read administrative boundaries data for Cambodia from a GeoPackage file and transform it to a specific CRS
Cambodia_admin <- st_read("kontur_KHboundaries_KH_20230628.gpkg") %>% st_transform(3106)
```

#### **4. Check and Create Boundaries**:

Here we've inspect the 'name_en' column and create the boundary for
Cambodia.

```{r}
# Extract distinct values of a specific attribute from an 'sf' object
# Extract distinct values of the 'name_en' attribute (administrative region names in English) from the Cambodia_admin object
distinct_names <- Cambodia_admin %>% distinct(name_en)

# Print the distinct names to the console
print(distinct_names)
```

***distinct_names:*** This is the name given to the resulting data frame
containing the distinct names extracted from the 'name_en' attribute of
the *Cambodia_admin* object.

By using these lines of code, we're extracting distinct administrative
region names in English from the Cambodia_admin object and printing them
to the console. This can be useful for exploring the unique values
present in the 'name_en' attribute and understanding the data better.

```{r}
# Creating Cambodia Boundary

# Extract the geometries (shapes) from the Cambodia_admin object
# Merge (union) all geometries into a single geometry representing the boundary of Cambodia
# Convert the merged geometry into an 'sf' object
# Ensure that the geometry is valid (i.e., without self-intersections or other topological errors)

Cambodia_boundary <- Cambodia_admin %>%
  st_geometry() %>%
  st_union() %>%
  st_sf() %>%
  st_make_valid()
```

**Explanation**:

**st_geometry()**: This function extracts the geometries (shapes) from
the 'sf' object Cambodia_admin It ensures that we are working with the
geometries themselves, rather than the entire data frame.

**st_union()**: This function merges (unions) all geometries into a
single geometry representing the boundary of Cambodia It effectively
creates a single polygon that encompasses all the individual polygons in
the Cambodia_admin object.

**st_sf()**: This function converts the merged geometry into an 'sf'
object. It ensures that the resulting object retains the spatial
attributes and metadata associated with the original object.

**st_make_valid():** This function ensures that the geometry is valid by
fixing any self-intersections or other topological errors. It returns a
valid geometry, which is important for subsequent spatial operations and
analyses.

**Cambodia_boundary:** This is the name given to the resulting 'sf'
object representing the boundary of Cambodia It contains a single
polygon geometry that encompasses all the administrative boundaries
present in the Cambodia_admin object.

By using these lines of code, we're creating a boundary for Cambodia by
merging all the administrative boundaries present in the Cambodia_admin
object into a single polygon. This boundary can then be used for various
spatial analyses and visualizations.

#### **5. Plot Boundaries for Verification**:

Visualize the hex data and boundaries to ensure accuracy.

```{r}
names(Cambodia_hex)
```

```{r}
require(plotly)

# Create the ggplot object
# This line initializes a ggplot object, which serves as a canvas for plotting.
# It starts with plotting the spatial data stored in the Cambodia_hex object.

p <- ggplot(Cambodia_hex) + 
  # Add polygons representing population data to the plot
  geom_sf(aes(fill = population), color = "gray66", linewidth = 1) + 
  # Add the boundary of Cambodia as a separate layer without fill
  geom_sf(data = Cambodia_boundary, fill = NA, color = "black")

# Convert ggplot object to a plotly object
# This line converts the ggplot object 'p' into an interactive plotly object.
# The 'tooltip' argument specifies that the 'population' attribute will be displayed when hovering over data points.

p <- ggplotly(p, tooltip = c("population"))

# Print the plotly object
# This line prints the interactive plotly object to the console or the viewer pane, allowing for interactive exploration.

print(p)
```

Explanation:

In this work, we made sure to install the ***plotly*** package to enable
interactive plotting capabilities. Then, we created a ggplot object as
our base canvas for plotting. We started by plotting the spatial data
stored in the *Cambodia_hex* object, adding polygons representing
population data to the plot, and including the boundary of Cambodia as a
separate layer without fill. After that, we converted the ggplot object
'p' into an interactive plotly object. We specified that the
'[*population*]{.underline}' attribute will be displayed when hovering
over data points. Finally, we printed the interactive plotly object 'p'
to the console or the viewer pane, allowing for interactive exploration
and sharing with others in our team.

#### **6. Calculate Aspect Ratio**:

Determine the aspect ratio for the map based on the bounding box of the
boundary.

```{r}
# Setting the bounding box of the plot to match the boundary of Cambodia
# We calculate the bounding box using the st_bbox() function, which returns the minimum and maximum coordinates of the Cambodia_boundary object.
bbox <- st_bbox(Cambodia_boundary)

# Finding the aspect ratio of the plot
# We create four points at the corners of the bounding box to calculate the width and height.
# Each point represents a corner of the bounding box: bottom-left, bottom-right, top-left, and top-right.
# We ensure that each point is in the same coordinate reference system (CRS) as the boundary data.
bottom_left <- st_point(c(bbox[["xmin"]], bbox[["ymin"]])) %>%
  st_sfc(crs = 3106)
bottom_right <- st_point(c(bbox[["xmax"]], bbox[["ymin"]])) %>%
  st_sfc(crs = 3106)
top_left <- st_point(c(bbox[["xmin"]], bbox[["ymax"]])) %>%
  st_sfc(crs = 3106)
top_right <- st_point(c(bbox[["xmax"]], bbox[["ymax"]])) %>%
  st_sfc(crs = 3106)

# Calculate the width and height of the plot using the distances between the corner points.
width <- st_distance(bottom_left, bottom_right)
height <- st_distance(bottom_left, top_left)

# Determine the aspect ratio of the plot based on the width and height.
# If the width is greater than the height, we set the width ratio to 1 and calculate the height ratio.
# Otherwise, we set the height ratio to 1 and calculate the width ratio.
if (width > height) {
  w_ratio = 1
  h_ratio = height / width
} else {
  h_ratio = 1.1
  w_ratio = width / height
}
```

**Explanation:**

Setting the bounding box: We use the st_bbox() function to calculate the
bounding box of the Cambodia_boundary object. This bounding box defines
the extent of the plot.

Finding the aspect ratio: We create four points at the corners of the
bounding box to calculate the width and height of the plot. Each point
represents a corner of the bounding box: bottom-left, bottom-right,
top-left, and top-right. We ensure that each point is in the same
coordinate reference system (CRS) as the boundary data.

Calculating width and height: We calculate the distances between the
corner points to determine the width and height of the plot.

Determining aspect ratio: Based on the width and height, we determine
the aspect ratio of the plot. If the width is greater than the height,
we set the width ratio to 1 and calculate the height ratio accordingly.
Otherwise, we set the height ratio to 1.1 (arbitrarily chosen for better
visualization) and calculate the width ratio accordingly.

#### 7. **Rasterize Population Data**:

```{r}
# Convert spatial polygons to raster format, then convert to matrix
# We determine the size of the output raster by multiplying the desired size by the width and height ratios.
# This ensures that the raster has the correct aspect ratio to match the plot.
size = 1000 * 3.5

# Convert spatial polygons to raster format
# We use the st_rasterize() function to rasterize the Cambodia_hex polygons.
# The 'nx' and 'ny' arguments determine the number of cells in the x and y dimensions of the raster.
# We calculate these values based on the desired size of the raster and the width and height ratios calculated earlier.
pop_raster <- st_rasterize(
  Cambodia_hex,
  nx = floor(size * w_ratio) %>% as.numeric(),
  ny = floor(size * h_ratio) %>% as.numeric()
)

# Convert raster to matrix
# We extract the 'population' values from the raster and arrange them into a matrix.
# The 'nrow' and 'ncol' arguments specify the number of rows and columns of the matrix, based on the size and aspect ratio of the raster.
pop_matrix <- matrix(pop_raster$population,
                     nrow = floor(size * w_ratio),
                     ncol = floor(size * h_ratio))

```

#### **8. Define Color Palette**:

```{r}
# Create color palette from the MetBrewer library
# We use the met.brewer() function from the MetBrewer package to generate a color palette named "Peru2".
# This palette is designed specifically for thematic mapping and provides a range of colors suitable for visualizing spatial data.
color <- MetBrewer::met.brewer(name="OKeeffe1", direction = -1)

# Generate color gradients from the color palette
# We use the colorRampPalette() function from the grDevices package to generate a smooth color gradient from the specified color palette.
# The 'bias' parameter adjusts the color gradient by emphasizing certain colors in the palette.
# We generate 256 colors to ensure a smooth transition between colors in the gradient.
tx <- grDevices::colorRampPalette(color, bias = 4.5)(256)

# Display color swatches for the generated colors
# We use the swatchplot() function from the MetBrewer package to visualize the color palette.
# This function displays a grid of color swatches representing the colors in the palette.
swatchplot(tx)

# Display color swatches for the original palette
# We also display color swatches for the original palette obtained from the MetBrewer library.
# This allows us to compare the generated color gradient with the original palette.
swatchplot(color)

```

#### **9. Render 3D Map**:

Use Rayshader to create a 3D representation of the population density.

```{r}
# Close any existing 3D plot before plotting another
# We make sure to close any existing 3D plot to start with a clean slate.
rgl::close3d()

# Apply height shading to the population matrix using the provided texture
# We apply height shading to enhance the visual depth of the population matrix, making terrain features more pronounced.
# The 'texture' argument specifies the color palette (tx) to be applied to the height shading.
pop_matrix %>%
  height_shade(texture = tx) %>%

# Plot the 3D visualization of the population matrix
# We create a 3D plot of the population matrix using the plot_3d() function.
# Adjusting 'zscale', 'solid', and 'shadowdepth' will affect the appearance of the plot.
# Higher 'zscale' values exaggerate the vertical relief, while lower values flatten it.
# Setting 'solid' to FALSE creates a wireframe representation, allowing us to see through the surface.
# Adjusting 'shadowdepth' changes the intensity of shadows cast by objects in the plot.
plot_3d(heightmap = pop_matrix,
          zscale = 250 / 4.5,   # Higher zscale exaggerates vertical relief 100/2 or 200/6
          solid = FALSE,        # Wireframe representation allows seeing through the surface
          shadowdepth = 0.8    # Adjusts intensity of shadows cast by objects
          )

# Adjust the camera angle for better visualization
# We adjust the camera angle to optimize the viewing perspective of the 3D plot.
# 'theta' controls the horizontal viewing angle, 'phi' controls the vertical viewing angle, and 'zoom' adjusts the distance between the camera and the plot.
render_camera(theta = 0,      # Adjusts horizontal viewing angle
              phi = 70,       # Adjusts vertical viewing angle
              zoom = 0.8)     # Controls distance between camera and plot


# To interactively view the 3D plot
# We generate an interactive widget to allow team members to explore the 3D plot interactively.
# This enables features such as rotation, zooming, and panning for a more immersive experience.
rgl::rglwidget()
```

#### Explanation:

**theta**: The 'theta' parameter of the render_camera() function
controls the rotation around the vertical axis, allowing us to adjust
the horizontal viewing angle. A value of 0 would position the camera
facing directly towards the positive x-axis, with positive values
rotating the camera clockwise and negative values rotating it
counterclockwise.

**phi**: The 'phi' parameter controls the elevation angle, allowing us
to adjust the vertical viewing angle. A value of 0 would position the
camera parallel to the xy-plane, looking straight down, with positive
values tilting the camera upwards and negative values tilting it
downwards.

**zoom**: The 'zoom' parameter controls the distance between the camera
and the plot. A value greater than 1 would zoom in, bringing the plot
closer to the camera, while a value less than 1 would zoom out, moving
the plot away from the camera. We set it into one for a balance look.

**zscale**: The 'zscale' parameter of the plot_3d() function adjusts the
scaling factor for the z-axis, determining the vertical exaggeration of
the plot. A higher value would exaggerate the vertical relief, while a
lower value would flatten it. We set it into 25 that is 200 / 8.

**solid**: The 'solid' parameter of the plot_3d() function controls
whether the surface should be plotted as a solid (filled) object or as a
wireframe (empty). Setting it to FALSE results in a wireframe
representation, allowing us to see through the surface to the underlying
structure.

**shadowdepth:** The 'shadowdepth' parameter of the plot_3d() function
controls the intensity of shadows cast by the objects in the plot. A
higher value results in deeper and darker shadows, while a lower value
softens the shadows. In this work, we set it into 0.8 for a softer
shadow.

#### **10. Render in high-quality and Save Image**:

Fine-tune the camera angle and render a high-quality image of the 3D
map.

```{r}
outfile <- glue::glue("KH_final_4.png")

{
  start_time <- Sys.time()
  cat(crayon::cyan(start_time), "\n")
  if(!file.exists(outfile)) {
    png::writePNG(matrix(1), target = outfile)
  }
  
 # Render a high-quality image of the plot
render_highquality(
    filename = outfile,               # Specify the output filename for the high-quality image
    interactive = FALSE,              # Disable interactivity for the rendered image
    lightdirection = 55,          # Set the direction of the primary light source (in degrees)
    lightaltitude = c(30, 80),        # Set the altitude of both lights (in degrees)
    lightcolor = c("white",'white'),    # Set the color of both lights (custom color and white)
    lightintensity = c(600, 100),     # Set the intensity of both lights
    width = 3500,                     # Set the width of the rendered image
    height = 3500,                    # Set the height of the rendered image
    samples = 256                     # Set the number of samples for anti-aliasing #250
)

  
  end_time <- Sys.time()
  diff <- end_time - start_time
  cat(crayon::cyan(diff), "\n")
}
```

\newpage

#### **11. Annotate the image**

You can add names and more details about your generated visualization.

```{r}
library(showtext)
library(extrafont)

pop_raster <- image_read("KH_final_4.png")

text_color <- "#003C84"
swatchplot(text_color)


# Automatically enable font support
showtext_auto()

# Download and register the Philosopher font from Google Fonts
font_add_google("Philosopher", regular = "400", bold = "700")

pop_raster %>%
  image_annotate("Cambodia",
                 gravity = "northeast",
                 location = "+50+50",
                 color = text_color,
                 size = 300,
                 font = "Philosopher",
                 weight = 700,

  ) %>%
  image_annotate("POPULATION DENSITY MAP",
                 gravity = "northeast",
                 location = "+50+340",
                 color = text_color,
                 size = 100,
                 font = "Philosopher",  # Corrected font name
                 weight = 500,

  ) %>%
  image_annotate("Visualization by: Babac, Gier, Laspoña, Mugot \nData: Kontur Population 2023",
                 gravity = "southwest",
                 location = "+20+20",
                 color = alpha(text_color, .8),
                 font = "Philosopher",  # Corrected font name
                 size = 65,

  ) %>%
  image_write("Plots/Annotated_plot_KH.png", format = "png", quality = 100)
```

## **Detailed Analysis Report: 3D Population Density Visualization of Cambodia**

**Introduction**:

In this comprehensive analysis, we delve into the creation and
interpretation of a 3D visualization representing the population density
across Cambodia. By leveraging spatial data and advanced visualization
techniques, we aim to provide in-depth insights into population
distribution patterns across different regions of the country.

**Methodology**: Data Acquisition and Preparation

Spatial data for Cambodia, including population data and administrative
boundary information, was sourced from GeoPackage files. The data
underwent preprocessing, including transformation to a specific
Coordinate Reference System (CRS) and extraction of relevant attributes
such as population and administrative region names.

**Visualization Techniques:**

The worked utilized the ***rayshader*** package in R, a powerful tool
for creating 3D visualizations, to generate a plot of the population
density. The visualization involved the conversion of spatial polygons
to raster format, followed by the application of height shading to
enhance visual depth.

**Color Mapping and Representation:**

A color palette from the ***MetBrewer*** library was selected to
represent population density gradients. The color palette provided a
visually appealing representation of population density variations, with
distinct colors indicating different population levels.

### **Results**:

The 3D visualization effectively portrays variations in population
density across Cambodia. Areas with higher population density are
depicted as elevated terrain, while areas with lower population density
appear as flat terrain.

In identifying high density areas, **Phnom Penh** emerges prominently as
the area with the highest population density. This is consistent with
its status as the capital city and economic center of Cambodia.
Provinces such as [***Kampong Cham, Kandal, Siem Reap***]{.underline},
and [***Battambang***]{.underline} also exhibit significant population
density, albeit lower than Phnom Penh.

**Discussion**:

Phnom Penh.

As the capital city, Phnom Penh serves as the political, economic, and
cultural hub of Cambodia. The high population density can be attributed
to factors such as employment opportunities, educational institutions,
healthcare facilities, and infrastructure development. Kampong Cham,
Kandal, Siem Reap, and Battambang:

These provinces represent some of the most populous regions in Cambodia
after Phnom Penh. Kampong Cham and Kandal benefit from their proximity
to Phnom Penh, contributing to their relatively high population density.
Siem Reap, renowned for the Angkor Wat temple complex, attracts tourists
and residents alike, driving population growth. Battambang, located in
the northwest region, serves as an agricultural and commercial center,
supporting population growth and urban development.

**Conclusion:**

The detailed analysis of the 3D population density visualization
provides valuable insights into regional demographic patterns in
Cambodia. By identifying areas of high population density such as Phnom
Penh, Kampong Cham, Kandal, Siem Reap, and Battambang, stakeholders can
make informed decisions regarding urban planning, infrastructure
development, resource allocation, and socioeconomic policies. This
analysis underscores the significance of spatial data visualization
techniques in understanding population dynamics and facilitating
evidence-based decision-making for sustainable development and inclusive
growth.

End.
